The Name of the Title Is Hope

### 2025-01-19 paper.tex
Cooking

Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available when a workgroup begins its lookback. This uncertainty arises because workgroups are scheduled independently \thomas{Possibly confusing. . . Not in lockstep?}, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction. In such cases, the dependent workgroup enters a waiting state, spinning until the reduction becomes available. During this waiting period, the spinning workgroup remains scheduled, consuming computational resources. Without \emph{forward-progress guarantees}, the scheduler may schedule the spinning workgroup indefinitely over its predecessor, preventing the predecessor from completing its reduction. This satisfies condition 4b, creating a circular dependency that results in deadlock within the kernel.

A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time. Sorensen et al.~\cite{barriers, 1, 2} provide the most comprehensive analysis of GPU schedulers. However, while they classify \emph{Chained-Scan} as an FPG-reliant algorithm, their primary focus is in exploring and formalizing progress models, rather than their application to any one algorithm. In the context of \emph{Chained-Scan} deadlocking, FPG prevents condition 4b by prohibiting indefinite scheduling of any single subgroup, thereby breaking the circular dependency chain.
\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct. The definition of HSA is \newline "Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated" \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA? It seems that it should be classified under OBE.}

\thomas{This example may be redundant now.}
For example, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor. Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG\@. In this setup, both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, loading the value at the memory address for the reduction of \emph{Work Tile 0} with each iteration. However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---its absence on key segments of hardware is a critical issue for developers. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\thomas{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery.} and at worst, risks the aforementioned TDR and subsequent program crash.

@@

\subsection{Earlier Attempts at Portability}
Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress requirements in \emph{ScalarFallback}. By allowing a workgroup to access and operate on another tile’s data, \emph{ScalarFallback} targets the mutual exclusion condition outlined by Coffman et al., ensuring that no single workgroup holds exclusive control over a resource. Instead, each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time. Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because a workgroup can perform the reduction of a blocking tile itself, breaking the inter-workgroup dependency.

@@

\section{Design}
\thomas{I've broken somewhat with the outline and moved the goals and non-goals section to somewhere in the introduction. I kept finding myself trying to write around it. I think the content there is good, but it's very awkward as it ended up so deep within the paper. The placement of this section here feels more natural, and seems to mirror other papers?}

\subsection{Breaking Deadlocks}
\thomas{What I want to accomplish here is to go beyond saying "Here is our method. It works great." I want to show that these are the possible avenues to attack deadlocking . . . and exhaustively demonstrate that removing  Coffman's mutual exclusion is the best way to do it, under our given constraints.}

Deadlock in \emph{Chained-Scan} arises from the inter-workgroup dependencies that are inherent to scan operations. These dependencies are governed by Coffman et al.'s~\cite{10.1145/356586.356588} four necessary conditions for deadlock: \emph{mutual exclusion}, \emph{resource holding}, \emph{no preemption}, and \emph{circular dependency}. While all four conditions are theoretically avenues for addressing deadlock, practical constraints and the requirements of \emph{Chained-Scan} significantly narrow the set of viable approaches.

The first condition, \emph{dependency relationships}, is immutable in the context of scan operations. Similarly, \emph{no preemption} is immutable within the GPU programming model. Addressing \emph{resource holding} by allowing workgroups to release resources (losing context) during waits effectively reintroduces kernel launches as inter-workgroup barriers. Designing tile sizes to break \emph{circular dependency} by mapping workgroups one-to-one with multiprocessors might seem feasible but is impractical. GPU vendors provide no guarantees regarding workgroup progress behavior, leaving starvation risks unresolved. Moreover, such strategies are non-portable and contradict the goal of supporting diverse hardware architectures.

\subsection{Decoupled-Fallback}

\section{Implementation}
We are in WGSL. We don't have: a memory model, 64-bit atomics, and atomicCAS is unreliable.

\subsection{Split Method}
\begin{enumerate}
    \item Idea: Split the reductions into multiple values; use separate values for local/inclusive instead of using flags to differentiate them.
    \item Pro: Full 32 or theoretically 64 bits.
    \item Pro: No memory model or 64 bit atomics required!
    \item Pro: Can safely fallback insert without risking clobbering.
    \item Con: 2x Memory usage
    \item Con: Places subgroup functions adjacent to atomics, which may lead to \textbf{\textit{interesting}} behavior.
\end{enumerate}

\subsection{Intra-Workgroup Scan}
Our intra-workgroup implementation is most similar to the \emph{CUDPP}~\cite{} scan kernel, with adaptations to enhance portability. It consists of three phases:
\begin{enumerate}
  \item
\end{enumerate}

The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer scan network embedded with Kogge-Stone subgroup scans to perform the scan on the intra-workgroup spine. The upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction; assuming shared memory bank widths equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a broadcast operation, resulting in no conflicts. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes.

\begin{algorithm}[htbp]
  \SetAlgoLined
  \KwIn{Shared Memory Array to Scan $x$, Workgroup Size $W$, Subgroup Size $S$}
  \KwOut{Shared Memory Inclusive Scanned Array $x$}

@@

  \ForEach{$thread\_id$ \textbf{in} $W$ \textbf{in parallel}}{
    \For{$j \gets S$ \KwTo $alignment$ \textbf{with} $j \gets j * S$}{

@@

@@

        \If{$cond1 \textbf{ \&\& } cond2$}{
          $x[fanout\_index] \gets x[fanout\_index] + x[((fanout\_index / stride) + top\_offset) - 1]$\;
        }
      }
      $stride \gets stride * S$\;
      $top\_offset \gets top\_offset + step$\;
    }
    \textbf{barrier()}\;
  }
  \Return{$x$}\;
  \caption{Subgroup-Size-Agnostic Scan}
  \label{alg:example}
\end{algorithm}

@@

\section{Experimental Setup}

\section{Results and Analysis}

@@

>>>>

Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available when a workgroup begins its lookback. This uncertainty arises because workgroups are scheduled individually \thomas{Possibly confusing. . . and perhaps not even true? Not in lockstep?}, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction. In such cases, the dependent workgroup enters a waiting state, spinning until the reduction becomes available. During this waiting period, the spinning workgroup remains scheduled, consuming computational resources. Without \emph{forward-progress guarantees}, the scheduler may schedule the spinning workgroup indefinitely over its predecessor, preventing the predecessor from completing its reduction. This satisfies condition 4b, creating a circular dependency that results in deadlock within the kernel.

A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time. Sorensen et al.~\cite{barriers, 1, 2} provide the most comprehensive analysis of GPU schedulers. However, while they classify \emph{Chained-Scan} as an FPG-reliant algorithm, their primary focus is in exploring and formalizing progress models, rather than their application to any one algorithm. In the context of \emph{Chained-Scan} deadlocking, FPG prevents condition 4b by prohibiting indefinite scheduling of any single subgroup, thereby breaking the circular dependency chain.
\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct. The definition of HSA is \newline "Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated" \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA? It seems that it should be classified under OBE\@.}

\thomas{This example feels redundant now. Timeout detection and recovery can be moved into the 2nd paragraph in this section.}
For example, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor. Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG\@. In this setup, both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, loading the value at the memory address for the reduction of \emph{Work Tile 0} with each iteration. However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---its absence in key segments of hardware is a critical issue for developers. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\thomas{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery.} and at worst, risks the aforementioned TDR and subsequent program crash.

@@

\subsection{Earlier Attempts at Portability}
Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress requirements in \emph{ScalarFallback}. By allowing a workgroup to access and operate on another tile’s data, \emph{ScalarFallback} targets Coffman et al.'s \emph{mutual exclusion} condition. Instead, each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time. Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because each workgroup can perform the reduction of any blocking tile, breaking the inter-workgroup dependency.

@@

\section{Decoupled-Fallback}
\thomas{I've broken somewhat with the outline and moved the goals and non-goals section to somewhere in the introduction. I kept finding myself trying to write around it. I think the content there is good, but it's very awkward as it ended up so deep within the paper. The placement of this section here feels more natural, and seems to mirror other papers?}
\emph{Decoupled-Fallback} enhances the \emph{Scalar-Fallback} technique with significantly improved fallback performance. As an extension of the \emph{Chained-Scan} architecture, it preserves the advantages in speed and oversubscription, making it an ideal choice for our requirements.

\thomas{What I want to accomplish here is to go beyond saying "Here is our method. It works great." I want to show that these are the possible avenues to attack deadlocking . . . and exhaustively demonstrate that removing mutual exclusion is the best way to do it, under our given constraints.}
Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to break deadlocking. \emph{Dependency relationships} and \emph{no preemption} are immutable characteristics of scan and the GPU programming model, eliminating them as viable avenues for breaking deadlocking. While it may be possible to break \emph{circular ordering} by designing tile sizes such that workgroups are mapped one-to-one to multiprocessors, doing so portably is infeasible. Moreover, vendor guidance on progress guarantees among multiprocessors is also lacking, leaving the same risks of deadlocking but at the multiprocessor level. While eliminating \emph{resource holding}---allowing workgroups to discard work tiles when they encounter a deadlock---is theoretically possible, it is strictly worse than eliminating \emph{mutual exclusion}, as it necessitates $t$ additional movements (discard-read-read versus read).

To maximize the efficiency of the fallback operation, we employ three strategies. First, although a lookback is performed by a single thread or subgroup, we want all threads in the workgroup to participate. To facilitate this, we use a workgroup-wide lock to allow threads participating in the lookback to signal the entire workgroup when a fallback is necessary. Second, we explicitly limit the number of times a workgroup may spin while waiting on a tile. If the maximum spin count is exceeded, the fallback operation is initiated. Although this approach still risks performing redundant fallbacks---since a work tile that is late is indistinguishable from one that is deadlocking---exposing the maximum spin count as a tunable parameter allows us to reduce the likelihood of false-positive fallbacks. Finally, after completing a fallback, we attempt to insert the value into the spine at the position of the deadlocking tile, preventing succeeding workgroups from initiating redundant fallbacks.

The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}. Regardless of FPG, due to false positives, any given hardware will have a natural fallback-incidence rate $f$. Consequently, we expect memory bandwidth traffic to increase (and throughput to decrease) by $O(fn)$ to $O((2 + f)n)$. However, just as serial ordering guarantees that memory accesses to the spine fall within the $o$-width sliding window, memory accesses to the scan input are similarly guaranteed to fall within an $ot$-width sliding window, making them also highly likely to reside within some level of the cache. As we later empirically demonstrate, this cache residency significantly reduces the cost of memory access, such that the fallback operation is almost always bound by the compute required to reduce the blocking tile. As a pure reduction incurs very lower overhead, we observe throughput approaching $O(2n)$ on most devices.

\section{Implementation}
At a high level, our algorithmn follows a five phase construction:
\begin{enumerate} 
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$. 
  \item \textbf{Subgroup-Rake}: Each thread performs a private, serial scan across its vectors and then participates in $p$ raking subgroup scans. This allows the entire scan across the subgroup-block to be barrier-free and remain in registers. The final subgroup-block reduction is then posted to shared memory. 
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup reductions, writing the results back to shared memory upon completion. 
  \item \textbf{Decoupled-Fallback}: The first subgroup in the workgroup performs a lookback operation, engaging in fallback operations if necessary. Once completed, it posts the results to shared memory for the workgroup to consume. 
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output. 
\end{enumerate}
Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. These phases are most similar to the \emph{CUDPP}~\cite{} scan kernel, with adaptations to enhance portability. Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies.

\subsection{Intra-Workgroup Portability Adaptations}
As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations. 
\subsubsection{Subgroup-Raking}
We desire the smallest intra-workgroup spine possible. Not only does this reduce shared memory usage, but it also directly impacts the number of workgroup barriers encountered during the workgroup-wide scan, as this number is proportional to the size of the local spine. While large subgroups are relatively insensitive to this effect, small subgroups are particularly affected (reference depth here). To address this, we perform $p$ raking iterations of subgroup Kogge-Stone scans, where the highest-ranked thread passes its output into each succeeding scan. For a workgroup of size $w$, this approach reduces the number of reductions that each subgroup must post from $p$ to $1$, and it decreases the size of the spine from $\frac{pw}{s}$ to $\frac{w}{s}$.
\begin{algorithm}[htbp]
  \small
  \SetAlgoLined
  \KwIn{Shared Memory Array to Scan $x$, Workgroup Size $W$, Subgroup Size $S$}
  \KwOut{Shared Memory Inclusive Scanned Array $x$}

@@

  \ForEach{$thread\_id$ \textbf{in} $W$ \textbf{in parallel}}{
    \For{$j \gets S$ \KwTo $alignment$ \textbf{with} $j \gets j * S$}{
      \tcp{Brent-Kung Upsweep}

@@

      \tcp{Fanout}

@@

        \If{$cond1 \textbf{ \&\& } cond2$}{
          $x[fanout\_index] \gets x[fanout\_index] + x[((fanout\_index / stride) + top\_offset) - 1]$\;
        }
      }
      $stride \gets stride * S$\;
      $top\_offset \gets top\_offset + step$\;
    }
    \textbf{barrier()}\;
  }
  \Return{$x$}\;
  \caption{Workgroup-Wide Scan}
  \label{alg:example}
\end{algorithm}
\subsubsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer scan network embedded with Kogge-Stone subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a broadcast operation, resulting in no conflicts. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes. \footnote{Since we also require a workgroup-wide pure reduce operation, this can be achieved by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}

\subsection{Decoupled-Fallback}
To reiterate, the goal of a lookback is twofold: to determine the reduction of all preceding work tiles and to communicate those values to succeeding tiles to accelerate their own lookback operations. Following the \emph{Decoupled-Lookback} technique, we model the spine as a finite state machine, where each work tile comprising the spine may exist in one of three states:
\begin{itemize}
    \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
    \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
    \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
\end{itemize}
Each lookback is, therefore, a traversal from the \emph{Ready} state to the \emph{Inclusive} state. To illustrate this process, we present a flowchart (SEE LOOKBACK FLOWCHART). Prior to the kernel launch, all tiles in the spine are initialized to the \emph{Not Ready} state. As shown in the dotted section of the flowchart, fallbacks occur when a workgroup encounters a blocking tile and the maximum spin count is exceeded.

\subsubsection{Tile State Implementation}
To maintain a coherent view of the tile's state, two bits are required, corresponding to the three possible states described earlier. In \emph{Decoupled-Lookback}, Merrill and Garland propose two strategies for communicating tile state within a 32-bit value: Use two separate 32-bit values—one for the tile state and the other for the reduction value—combined with a \emph{memory fence}, ensuring that updating the tile state also synchronizes the reduction value. Use a single 64-bit value by bit-packing the tile state and reduction value, with updates managed via 64-bit atomics.

In our case, neither are viable options. Due to a lack of support across backend APIs, WebGPU does not furnish developers with a memory model, and currently, 64-bit atomics are not supported. Moreover, 64-bit atomics are not supported across most mobile devices, and thus we do not consider their future support an appropriate option for our implementation. Instead, we use two 32-bit values and split the reduction into upper and lower 16-bit segments, reserving the two most significant bits of each segment to encode the tile state.
\newline
SPLIT BITS MINI DIAGRAM here
\newline

\begin{enumerate}
  \item Idea: Split the reductions into multiple values; use separate values for local/inclusive instead of using flags to differentiate them.
  \item Pro: Full 32 or theoretically 64 bits.
  \item Pro: No memory model or 64 bit atomics required!
  \item Pro: Can safely fallback insert without risking clobbering.
  \item Con: 2x Memory usage
  \item Con: Places subgroup functions adjacent to atomics, which may lead to \textbf{\textit{interesting}} behavior.
\end{enumerate}

\begin{algorithm}[htbp]
  \tiny
  \SetAlgoLined
  \DontPrintSemicolon

  \KwIn{%
    \emph{Partition ID} $part\_id$, 
    \emph{Number of lanes} $lane\_count$, 
    \emph{Workgroup-shared lock} $wg\_lock$, 
    \emph{Atomic reduction array} \texttt{reduction}[], 
    \emph{Local partial reductions} $wg\_reduce[]$, 
    \emph{Constants:} $MAX\_SPIN\_COUNT, FLAG\_MASK, FLAG\_NOT\_READY, FLAG\_INCLUSIVE, FLAG\_READY, ALL\_READY, \dots$
  }
  \KwOut{Possible updates to $prev\_red$ or partial progress before fallback.}

  \If(\tcp*[h]{Only needed if not the first partition}){$part\_id \neq 0$}{
    $\,prev\_red \gets 0$\;
    $\,lookback\_id \gets part\_id - 1$\;
    $\,lock \gets \texttt{workgroupUniformLoad}(\&wg\_lock)$\;

    \While(\tcp*[h]{Spin until we can proceed or must fallback}){$lock = \textnormal{LOCKED}$}{
      \If(\tcp*[h]{Only threads within the subgroup size do checks}){$threadid.x < lane\_count$}{
        $\,spin\_count \gets 0$\;
        \While{$spin\_count < MAX\_SPIN\_COUNT$}{
          $\,flag\_payload \gets \texttt{select}\bigl(
             0,\;
             \texttt{atomicLoad}(\&\texttt{reduction}[\,lookback\_id\,][\,threadid.x\,]),\;
             threadid.x < \textnormal{SPLIT\_MEMBERS}
           \bigr)$\;

          \If{$\texttt{subgroupBallot}\bigl((\,flag\_payload \,\&\, FLAG\_MASK) > FLAG\_NOT\_READY\bigr).x = ALL\_READY$}{
            $\,incl\_bal \gets \texttt{subgroupBallot}\bigl(
               (\,flag\_payload \,\&\, FLAG\_MASK) = FLAG\_INCLUSIVE
             \bigr).x$\;

            \If{$incl\_bal \neq 0$}{
              \tcp{Wait for all lanes to show FLAG\_INCLUSIVE}
              \While{$incl\_bal \neq ALL\_READY$}{
                $\,flag\_payload \gets \texttt{select}\bigl(
                  0,\;
                  \texttt{atomicLoad}(\&\texttt{reduction}[\,lookback\_id\,][\,threadid.x\,]),\;
                  threadid.x < \textnormal{SPLIT\_MEMBERS}
                \bigr)$\;
                $\,incl\_bal \gets \texttt{subgroupBallot}\bigl(
                  (\,flag\_payload \,\&\, FLAG\_MASK) = FLAG\_INCLUSIVE
                \bigr).x$\;
              }
              $\,prev\_red \gets prev\_red + \texttt{join}\bigl(
                 flag\_payload \,\&\, \textnormal{VALUE\_MASK},\;threadid.x
               \bigr)$\;

              \If{$threadid.x < \textnormal{SPLIT\_MEMBERS}$}{
                $\,t \gets \texttt{split}\bigl(
                  prev\_red + wg\_reduce[\mathit{spine\_size}-1],\;threadid.x
                \bigr) \,\vert\, FLAG\_INCLUSIVE$\;
                $\texttt{atomicStore}\bigl(
                  \&\texttt{reduction}[\,part\_id\,][\,threadid.x\,],\; t
                \bigr)$\;
              }
              \If{$threadid.x = 0$}{
                $\,wg\_lock \gets \textnormal{UNLOCKED}$\;
                $\,wg\_broadcast \gets prev\_red$\;
              }
              \textbf{break}\;  % Break out of spin loop
            }
            \Else{
              $\,prev\_red \gets prev\_red + \texttt{join}\bigl(
                 flag\_payload \,\&\, \textnormal{VALUE\_MASK},\;threadid.x
               \bigr)$\;
              $\,spin\_count \gets 0$\;
              $\,lookback\_id \gets lookback\_id - 1$\;
            }
          }
          \Else{
            $\,spin\_count \gets spin\_count + 1$\;
          }
        }
        \If(\tcp*[h]{Escalate fallback if spinning too long}){%
          $\,threadid.x = 0 \,\wedge\, spin\_count = MAX\_SPIN\_COUNT$}{
          $\,wg\_broadcast \gets lookback\_id$\;
        }
      }
      \tcp{(At this point, if still locked, we consider fallback.)}
      $\,lock \gets \texttt{workgroupUniformLoad}(\&wg\_lock)$\;
      \textbf{break}\;
    }
  }
  \caption{“Lookback” portion of the single-subgroup scan.}
  \label{alg:lookback-portion}
\end{algorithm}

\begin{algorithm}[htbp]
  \tiny
  \SetAlgoLined
  \DontPrintSemicolon

  \KwIn{%
    \emph{Workgroup lock} $wg\_lock$, 
    \emph{Current broadcast index} $wg\_broadcast$, 
    \emph{Local partial} $prev\_red$, 
    \emph{Reduction array} \texttt{reduction}[], 
    \emph{Local partials} $wg\_reduce[]$, 
    \emph{Constants:} $FLAG\_MASK, FLAG\_READY, FLAG\_INCLUSIVE, \dots$
  }
  \KwOut{%
    Possibly unlocked $wg\_lock$ and updated partition's inclusive data
  }

  \tcp{Check if we're still locked and must fallback}
  \If{$\textnormal{lock} = \textnormal{LOCKED}$}{
    $\,fallback\_id \gets wg\_broadcast$\;

    \tcp{Invoke the fallback reduce subroutine}
    $\,f\_red \gets \textbf{FallbackReduce}\bigl(fallback\_id,\, s\_offset,\, \dots\bigr)$\;

    \If{$threadid.x < lane\_count$}{
      $\,f\_split \gets \texttt{split}(f\_red,\, threadid.x)\,\vert\,
        \texttt{select}(FLAG\_READY,\, FLAG\_INCLUSIVE,\,(fallback\_id=0))$\;

      \tcp{Atomically update if in SPLIT\_MEMBERS}
      $\,f\_payload \gets 0$\;
      \If{$threadid.x < \textnormal{SPLIT\_MEMBERS}$}{
        $\,f\_payload \gets \texttt{atomicMax}\bigl(
           \&\texttt{reduction}[fallback\_id][threadid.x],\, f\_split
         \bigr)$\;
      }

      $\,incl\_found \gets \Bigl(
         \texttt{subgroupBallot}\bigl((\,f\_payload \,\&\, FLAG\_MASK) = FLAG\_INCLUSIVE\bigr).x 
         = \textnormal{ALL\_READY}
       \Bigr)$\;

      \If{$incl\_found$}{
        $\,prev\_red \gets prev\_red + \texttt{join}\bigl(
           f\_payload \,\&\, \textnormal{VALUE\_MASK},\;threadid.x
         \bigr)$\;
      }
      \Else{
        $\,prev\_red \gets prev\_red + f\_red$\;
      }

      \If(\tcp*[h]{Either first partition or all splits are inclusive}){$(fallback\_id=0)\,\vee\,(incl\_found)$}{
        \If{$threadid.x < \textnormal{SPLIT\_MEMBERS}$}{
          $\,t \gets \texttt{split}\bigl(
             prev\_red + wg\_reduce[\mathit{spine\_size}-1],\;threadid.x
           \bigr)\,\vert\,FLAG\_INCLUSIVE$\;
          $\texttt{atomicStore}\Bigl(
             \&\texttt{reduction}[part\_id][threadid.x],\, t
           \Bigr)$\;
        }
        \If{$threadid.x = 0$}{
          $\,wg\_lock \gets \textnormal{UNLOCKED}$\;
          $\,wg\_broadcast \gets prev\_red$\;
        }
      }
      \Else{
        $\,lookback\_id \gets lookback\_id - 1$\;
      }
    }

    $\,lock \gets \texttt{workgroupUniformLoad}(\&wg\_lock)$\;
  }

  \caption{“Fallback” logic, calling \textbf{FallbackReduce} if still locked.}
  \label{alg:fallback-logic}
\end{algorithm}

@@

\section{Evaluation}
256 threads, 16 elements per thread. Shared memory requirements. RTS implementation shares an identical local implementation. Uses Merrill-Grimshaw style raking, with our own occupancy discovery technique.

\subsection{Comparison to Reduce-Then-Scan}

\subsection{How (un)fair is fair?}
Show the stats of each device. How many lookbacks?  How many fallbacks initiated? How many successful fallbacks?

\subsection{Emulated Deadlocking}
Show the performance of devices as we artificially increase the fallback-incidence rate.

@@

\thomas{Somewhere in here is perfect for Raph's commentary on the speed of atomics on different hardware. Once we have it, we can retool this section around it.}
